Seabird assemblages are linked to the major western boundary current off eastern Australia

### 2024-03-06 ms_PROOCE/ms_PROOCE_review1_response-to-reviewers.Rmd

---
title: ""
# author: "Author"
# date: ""
output: pdf_document
---
`r Sys.Date()` \newline

\begin{flushright}
\textit{Submission}: \textbf{PROOCE-D-23-00336} \newline
\textit{Seabird assemblages are linked to the major western boundary current off eastern Australia} \newline
\end{flushright}

---

\begin{flushright}
\underline{\textbf{Response to Reviewer's questions comments}} \newline
\end{flushright}

\textcolor{blue}{Most of the comments from Reviewer 2 in this section are also under their detailed comments. Therefore, those were omitted from this section and answered/commented below.} \newline

*Question 5*

**Reviewer 1**

*Line 8 if the Highlights are those that the study has uncovered, the first ‘highlight’, southward movement of the EAC, is not that which is determined in this study.*

\textcolor{blue}{That is right, thanks for noticing it. We have now removed this sentence from the Highlight session.} \newline

*Line 341 'While remote tracking studies are advancing our understanding on how individual seabirds use their environment at fine scales (Tremblay et al., 2009)', this study as an understanding of how much tracking studies advance the understanding of seabird interaction and their environment fails to look at more current papers. A good example (with referenced other examples within) is: Trevail et al (2023) Tracking seabird migration in the tropical Indian Ocean reveals basin-scale conservation need. <https://doi.org/10.1016/j.cub.2023.10.060>*

\textcolor{blue}{Thanks for suggesting Trevail et al. (2023), which we have included alongside Tremblay et al. (2009).} \newline

**Reviewer 2**

\textcolor{blue}{See comments below.}

---

\begin{flushright}
\underline{\textbf{Response to Reviewer's 2 detailed comments}} \newline
\end{flushright}

*This study uses Region of Common Profile mixture models to describe seasonal seabird distributions with respect to environmental parameters, using at-sea survey data and oceanographic, climatic and physiographic factors. The authors found two macro-scale seabird assemblages, defined as 'northern' and 'southern' assemblages separated roughly near the latitude where the East Australian Current (EAC) moves off the Australian continental slope (~34 S). The authors state that the two seabird assemblages were mostly persistent across four seasons. The authors suggest that the predicted poleward intensification of the EAC makes it imperative that changes to seabird distribution in this important region be monitored, using this study as a baseline.*

*This paper provides new information about an important seabird 'hot spot' in the South Pacific Ocean, and it is timely in that climate-driven changes to the region have implications to the seabird community. In addition, the results describe physical drivers that influence seabird distribution and assemblages in this marine system. The paper is generally well written and easy to follow.*

\textcolor{blue}{Thanks for your thorough, positive review of our manuscript. We greatly appreciate your time and dedication to commenting and suggesting points that could be clarified/better explored, which have undoubtedly improved the manuscript. Please find below replies to the main 'general aspects' you stressed out, followed by detailed replies for each specific topic.} \newline

*There are four general aspects for which I recommend clarification or changes.*

*First, unless this journal has a strict limit on length of the paper, I think the scope and integration of the various components of this study would be better served by moving some of the detail available in Supplemental material into the main paper. Or, perhaps as an attached appendix. In particular, I suggest including the results of species richness analysis, which seems relevant to the focus of the paper. I hope more about the assemblages themselves (species involved and what they are responding to) are covered in subsequent papers.*

\textcolor{blue}{Thanks for suggesting the inclusion of supplementary figures in the main paper. The species richness analysis was moved to the main text, while other figures, as explained below, were kept in the supplemental material.}

\textcolor{blue}{Also, thanks for encouraging a separate publication on the assemblages (a likely following paper on the way).} \newline

*Second, it's not clear why only the spring is presented for some results (Figures 2, 4, 5) in the main paper (except perhaps for brevity and simplicity). If the authors provided justification for why spring results were highlighted, I could not find it. Furthermore, Suppl. figures 9 and 10 appear to show considerable differences among seasons in the spatial distribution of the two assemblages. Although the Discussion also mentions the influence of summer/autumn migrants (lines 293-299), there are statements about the seasonal persistence of assemblages distributions, which is confusing. More explanation of the observed seasonal differences may be warranted.*

\textcolor{blue}{Although \textit{Progress in Oceanography} does not limit figures, we decided to keep supplementary figures as they were. We chose to present only spring results for the sake of the brevity of the main text. Readers interested in specific maps and graphs can refer to the supplementary material at their leisure. From our perspective, this does not influence the reader's understanding of the main messages of the paper.}

\textcolor{blue}{To clarify this in the main paper, we added the following sentence under the \textit{RCP models} heading:}

\textcolor{blue}{"In this section, we will present only the main results for the spring season when delving into specifics about each model and its results. The results of spring models summarise the key findings of our analysis, but we encourage the reader to look at the Supplementary Material for other seasons."}

\textcolor{blue}{We believe the differences in the spatial patterns of assemblages shown in the (original) supplementary figures 9 and 10 (predicted probability maps) are due to the lack of comprehensive sampling in all seasons. Given the strong support for a 'southern' and a 'northern' assemblage, we note that both summer and autumn fail to have observations in both southern/northern areas. The rarefaction curves also show this by lower values on 'sample coverage' for RCP1 (northern) in summer and RCP3 (northern) in autumn. Further, the point predictions assigned higher probabilities of 'southern' assemblage in the 'northern' area for summer and the opposite in autumn, suggesting some limitations about these results. We agree we could have explored these incongruences further in the main text, so we have now expanded this explanation on specific sections of the discussion, as detailed in the replies below.} \newline

*Third, I could not find any definition of the seasonal breakdowns - which months comprise summer, autumn, winter, spring? It may be clear to the authors, but studies vary widely in those seasonal categories. Also, please make it clear to northern readers that you are describing austral seasons.*

\textcolor{blue}{Thanks for noticing this. We have now added a sentence to include which months comprise each season (see below).} \newline

*Fourth, I suggest the authors make more effort to compare the patterns found in this study region with results from studies in other regions, including the northern hemisphere and sub-Arctic and Arctic.*

\textcolor{blue}{Thanks for your comment. This is the first study applying RCP models to highly mobile animals, more specifically to seabirds. Although there is an obvious component of seabirds throughout the paper, the key points in the discussion are broader--i.e. the application of RCP to highly mobile animals and marine biogeography. However, we believe these comparisons asked by the reviewer are among a sound background Intro section (L 29--43 and 50--56) and later subtly addressed in L 286--287 and 378--379 in the Discussion, according to the references provided. Further, the different oceanography dynamics, seabird families, and foraging strategies for the seabird species compared in a sub-Arctic/Arctic study with a tropical-cool temperate study do not seem adequate to us. Therefore, we would prefer not to add another paragraph to the main text.} \newline

*Specifics (by Line; L)*

*L 41-43: Consider including publications from other bioregions (some more recent), which also show SST and water masses as key drivers of seabird distributions. There are many; a few are: Gall et al. 2013, 2022; Nur et al. 2011, Serratosa et al. 2020.*

\textcolor{blue}{Thanks for the suggestion. We have included Serratosa et al. (2020) and Gall et al. (2022).} \newline

*L 90 - not sure a blanket statement like 'seabird distribution is under-represented in climate change studies' is the case elsewhere (e.g, studies in the North Atlantic, and to less extent in the North Pacific).*

\textcolor{blue}{Even though there are studies in the North Atlantic/Pacific, these are mainly around the USA; therefore, we believe seabirds are still under-represented in this field of research. However, we do acknowledge these efforts and studies. Thus, we have changed the sentence to reflect that the lack of these studies in the Australasia region is particularly noticeable.} \newline

*L 127-128 - Briefly describe the main features of this protocol, even though details can be obtained from the citation given. How were flying (vs on water or foraging) birds treated; how was distance (within 300m) determined; when were surveys halted due to conditions? Since 'one or more' observers recorded birds, how did you adjust for differences in effort?*

\textcolor{blue}{The protocol does not record distance from the ship for each individual record (i.e. it is a \textit{strip-transect} protocol); we have added this info in the section 2.2. We also added the behaviours, \textit{"(i.e. flying, feeding, sitting on the water)"}, in the section 2.2 and 2.3.2, as we summed all behaviours for our aggregated data set. In addition, we added the conditions when surveys were halted (\textit{"Seabird surveys were halted when sea conditions exceeded 6 m wave height and 40 kn of wind or visibility was under 500 m."}).}

\textcolor{blue}{Sorry, the issue raised about having 'one or more' observers was a misunderstanding between the authors. \textit{Two} observers continuously recorded seabirds throughout the day; single observers were only present during brief meal and comfort [toilet] breaks (5--15 min). So, there was no significant difference in effort, since the observations were confined to forward quarter 300 m and there was never the density of birds that reduced data entry or impacted on survey efficacy. This was fixed in the main text by changing from \textit{"One or more observers"} to \textit{"Two observers"}.} \newline

*L 137 - 'data wrangling' is jargon; revise to something like 'Data management', or 'compilation'.*

\textcolor{blue}{Data 'wrangling' is an actual concept in the field of Data Science, which means 'tidying and transforming' your data. See well-known data science book "R for Data Science", which has a whole section on 'wrangle' in its 1st edition (at \url{https://r4ds.had.co.nz/wrangle-intro.html}), and properly define the concept in its 2nd edition (at \url{https://r4ds.hadley.nz/intro}).} \newline

*Section 2.3.2 - somewhere in this section, describe what months constitute each season.*

\textcolor{blue}{Done. We have included a sentence that reads:}

\textcolor{blue}{"Seasons were defined as summer (Dec--Feb), autumn (Mar--May), winter (Jun--Aug), and spring (Sep--Nov)."} \newline

*L 187 - not clear what the 'species profiles' are here. Please clarify what this refers to.*

\textcolor{blue}{Done. We have ammended the sentence, which now reads:}

\textcolor{blue}{"RCPs can be directly interpreted as assemblages; the assemblage structure is given by the species profiles (i.e. the proportion of each species within the assemblage given by their probability of occurrence or predicted abundance)."} \newline

*L 231-233 - I think Supple Fig.2 (species richness maps) would be best in the main paper. While the rarefaction curves (Supple Fig. 13) serve to determine if sampling was adequate, the results in Fig. 2 seem directly relevant to the goal of describing the distributions of seabird assemblages.*

\textcolor{blue}{Thanks for pointing this out---we moved species richness analysis into the main paper.} \newline

*L 245-247 - authors note that in particular, 'proxies for persistent oceanographic features that may aggregate food…were selected only in the autumn and spring abundance models'. This is interesting, but I don't see in the Discussion where the authors propose why this may be. Somewhere in Discussion, it would also be helpful to indicate what prey are being affected, and where, by the oceanographic features and SST. Alternatively, the Methods section could include information on prey consumed by regional seabirds in the region.*

\textcolor{blue}{Thanks for your comment. We did expand on this interesting finding of only abundance models selecting variables summarising persistent oceanographic features (see the new sentences below). However, we think it is out-of-scope for our study to delve into prey and prey distribution. This would require new sections, from Methods to Discussion, and very likely new analyses, which may deserve a separate publication.}

\textcolor{blue}{The new sentences were added under section 4.2, splitting the (original) third paragraph into two. The sentences read as follows:}

\textcolor{blue}{"[...]. Although we have not approached the data in a multi-scale analysis, our results partially support this. Presence-absence models retained static variables and oceanographic variables that define water masses (i.e. SST and salinity). In addition to those, the autumn and spring abundance models also retained variables associated with persistent oceanographic features (e.g. climatic variables related to eddies and fronts), as well as mixed layer depth and chlorophyll in the autumn model. Eddies, fronts, high chlorophyll and shallower mixed layer depths are all related to increased seabird foraging success (Weimerskirch, 2007; Scales et al., 2014; De Pascalis et al., 2021), and therefore likely to influence seabird numbers. [...]"} \newline

*Fig. 5 - Might note in caption that the species along the Y-axis are arranged taxonomically.*

\textcolor{blue}{The species are ordered alphabetically (from bottom up) as the default output from the \{ecomix\} package. Nevertheless, we noted this in the figure caption to avoid doubts (in both the main paper and supplementary material).} \newline

*L 266-275 - Very nice opening paragraph. It summarizes key findings and implications to monitoring and conservation.*

\textcolor{blue}{Many thanks!} \newline

*L 277 - Suggest rewording to '..and water masses has been well established'. Could also include northern studies (e.g., Gall et al 2013 or 2022).*

\textcolor{blue}{Thanks, we have accepted your suggestion and included Gall et al. (2022) in the reference list.} \newline

*L 279 - as noted earlier, Supple Fig. 9 & 10 appear to show considerable spatial shifts in latitude, even if still two assemblages - thus the 'spatial consistency of these patterns throughout seasons' isn't obvious to me. They show similar patterns in winter and spring, but summer and autumn each have different patterns. A difference in pattern during autumn is mentioned in lines 254-255, but not addressed later. Please explain, or if you're basing this conclusion on another set of figures, cite those.*

\textcolor{blue}{We have included "(but see Section 4.3)" [limitations] after the L 279 sentence, and expanded on that in the proper section by writing a new paragraph:}

\textcolor{blue}{"These limitations are illustrated in the results for the summer and autumn seasons. Considering the strong support for a southern and a northern assemblage, we note that both summer and autumn lack samples in representative southern and northern (summer) and northern (autumn) areas. Albeit the sample coverage values were overall high, compared to winter and spring, summer and autumn had lower values (in particular, for the southern assemblage 'RCP1' in summer and for the northern assemblage 'RCP3' in autumn; Supplementary Fig. 12). Further, based on point predictions, higher probabilities of the southern assemblage in the northern area in summer (both presence-absence and abundance models) and of the northern assemblage in the southern area in autumn (abundance model), were assigned to a few grids (Fig. 4). The latter reinforces the possible weakness of these models to detect stronger relationships between the identified assemblages and the predictors. We also acknowledge the sharp spatial change in probabilities between assemblages in summer models (i.e. without a proper 'transition zone'), which may be related to the previous point. However, both summer and autumn models detected higher probabilities of expected taxa in their species profiles, suggesting that the make-up of the assemblage was well distinguished by the models. In addition, these models also seem to have captured the appropriate number of assemblages, agreeing with other models/seasons. Despite their limitations, we believe these models agree with the big picture of our results, therefore being a valid resource as a baseline."}

\textcolor{blue}{Regarding the 'three assemblage pattern' for autumn (L 254--255) mentioned by the reviewer, we believe we have touched on it in the second paragraph of section 4.1 and when concluding it by saying "[...] In turn, including these taxa in the models could result in a more stable 'three assemblage pattern' (as revealed by the autumn presence-absence model) in all but the summer season, when the majority of Antarctic species are breeding (Marchant and Higgins, 1990)."} \newline

*L 306-332 - I like this section describing the results in context of other descriptions of biogeographic units and LMEs. It's clearly written and is relevant to management issues.*

\textcolor{blue}{Thank you!} \newline

*L 340 - Suggest mentioning the positive aspects of studying seabirds in context of environmental parameters - i.e., seabirds are highly visible and therefore it's possible to obtain good sample sizes over large areas, and have high encounter rates in some locations (easy to identify hotspots). (there are multiple citations stating these points).*

\textcolor{blue}{We understand that for a broader readership, your suggestion might be valid (e.g. people not used to seabird surveys). Therefore, we have included the sentences below:}

\textcolor{blue}{"Among marine megafauna, seabirds are relatively easy to detect visually, enabling suitable sample sizes to be gathered over large areas (Ballance, 2007). Thus, seabirds can be used as models to study the distribution of highly mobile animals in relation to environmental characteristics in the marine environment."} \newline

*L365-367 - Could add some recent studies*

\textcolor{blue}{Done. We have added Commins et al. (2014) and Gall et al. (2022) in the first part of the sentence, and De Pascalis et al. (2021) on the second part of the sentece.} \newline

*L 373 - remove 'which we suspect do not exist….'. Replace with something like 'which to our knowledge are not generally available at present, ". You might check whether this is in fact the case, at least for your region.*

\textcolor{blue}{Thanks for the suggestion. To the best of our knowledge, no such data is available for our study area. So, we slightly modified the previous sentence to:}

\textcolor{blue}{"Such a hierarchical approach would require substantial concurrent observational and environmental data sets, which, to our knowledge, do not exist for our study area."} \newline

*L 381-384 - While I don't disagree with these statements on 'need for more seabird data', these are generally known concepts, and not directly relevant to scope of this journal. I would remove or reduce this paragraph to keep focus on the important findings.*

\textcolor{blue}{Thanks for your comment. The (original) paragraph was about our study's limitations and how the analyses could benefit from more data. As you stressed before (mainly regarding summer/autumn), this also has implications for the described patterns, which we have expanded the text on 'limitations' in the current version of the manuscript. Therefore we would prefer not to remove this paragraph.} \newline

*L 392 - Can authors provide information on why SST is a key variable discriminating the two seabird assemblages? The Discussion could provide brief examination of what SST is linked to: e.g,, the biological (prey type and distribution?) and physical (more on fronts and upwelling?) features birds are likely responding to. How do prey differ north and south of the 'transitional zone'?*

\textcolor{blue}{Our analysis does not allow us to dive into the 'whys' as a correlative method. However, as shown elsewhere, SST likely influences prey distribution, affecting predator distribution. We briefly mentioned that while presenting biogeographic schema that agreed with our findings, such as the mesopelagic biogeographic provinces from Sutton et al. (2017). As mentioned previously, we think it is out-of-scope to go further down the 'prey distribution' topic. Nonetheless, we have highlighted this in a new sentence in the Conclusions:}

\textcolor{blue}{"The mechanistic links between SST and seabird assemblages deserve more studies but may be due to the distribution of seabird prey, which at the biogeographic level seems to correlate with our findings (Sutton et al., 2017)."} \newline

*L 402 - might the 'transition zone' also shift poleward? Or might it expand in size? That would have implications for future seabird studies attempting to replicate your study.*

\textcolor{blue}{Good point. We believe the 'transition zone' might shift polewards following EAC's strengthening. We have modified the original sentence to reflect this; it now reads:} \newline

\textcolor{blue}{"In particular, the transition zone between assemblages might be pushed southwards along with the expected trend of the EAC. The transition zone between assemblages (Morrone, 2023), associated with the eastern extension of the EAC, may be fundamental to effectively monitoring the strengthening of the EAC and the subsequent impact on marine biota."} \newline

### 2024-03-06 ms_PROOCE/ms_PROOCE.Rmd
update README, update scripts-workflow figure, updated references inlcuding Daudt et al 2024 and 'corrected' refs for R pckgs, plus all source RMD files

authors:

@@

    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand

@@

    thanks: Corresponding author

@@

    department: Australasian Seabird Group
    affiliation: BirdLife Australia
    location: Hobart, Australia
    email: eric.woehler@gmail.com

@@

    department: Department of Mathematics and Statistics
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
    email: matthew.schofield@otago.ac.nz

@@

    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
    email: robert.smith@otago.ac.nz

@@

    department: Institute of Biological Sciences
    affiliation: Universidade Federal do Rio Grande---FURG
    location: Rio Grande, RS, Brazil
    email: lbugoni@yahoo.com.br

@@

    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
    email: will.rayment@otago.ac.nz

@@

  **Aim:** To identify assemblages of seabird species and relate them to environmental conditions off eastern Australia, a world seabird hotspot.
  **Location:** East coast of Australia.
  **Methods:** We used 15 seabird at-sea ship-based survey data sets collected over 37° of latitude, from 2016 to 2021, and fitted Region of Common Profile (RCP) mixture models for each season and type of data (presence-absence and abundance). RCP groups are defined as a set of regions where the probability of encountering a specific species profile is constant within regions, but different amongst them. These groups also vary according to covariates, which in our case included oceanographic, climatic, and physiographic parameters.
  **Results:** Results were based on 142,646 seabirds recorded, from 80 species, including albatrosses, petrels,  prions, shearwaters, boobies, and terns, among other taxa. All models suggested two macro-scale assemblages (‘northern’ and ‘southern’), except for the autumn presence-absence model that identified three groups. The model results consistently show a biogeographic transition at ~34°S, near the latitude at which the East Australian Current (EAC) separates from the Australian continental slope. Sea surface temperatures or sea surface salinities were selected in all final models, further indicating a close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and abundance, resulted in similar spatial and species profile patterns.
  **Main conclusions:** RCP models clearly identified two seabird assemblages off the east coast of Australia, suggesting the persistence of these groups at seasonal and macro spatial scales. Given ongoing poleward intensification that the EAC is experiencing, which is projected to continue over the next century, and its importance in influencing distributions of seabirds, the methods applied in our study could be replicated to assess possible changes in seabird assemblages, and how they are affected by changing environmental conditions.

@@

csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl

@@

    base_format: rticles::arxiv_article

@@

- \usepackage[left]{lineno}
- \newcommand{\blinenumbers}{\begin{linenumbers}}
- \newcommand{\elinenumbers}{\end{linenumbers}}

@@

@@

<!-- Note: I've changed the 'arxiv.sty' source file to include the running title in the PDF, and added commands in the YAML header too -->

```{r setup, include = FALSE, echo = FALSE}
library(readr)
library(dplyr)
library(purrr)
library(stringr)
library(kableExtra)

### If using {gt}:
# library(gt)
# options(gt.html_tag_check = FALSE)
```

# Biosketch {.unnumbered}

**Nicholas W. Daudt (he/him)** is broadly interested in spatial ecology and ecological modelling. His research uses data science and modelling to understand biogeographic patterns and ecological processes that build marine top predators' biodiversity patterns across space and time.

------------------------------------------------------------------------

**Data availability statement:** The gridded data used as the basis for all analyses are available in the archived repository along with the code to reproduce the results. The seabird raw data are currently being used for other projects and thus can not be publicly available. If you are interested in the seabird raw data, please contact Eric J. Woehler. The environmental data used in this study are freely available from the source (see Table 1 in the main text).

**Funding statement:** Parks Australia provided funding support to participants for selected voyages.

**Conflict of interest disclosure:** The authors declare no conflict of interest.

**Ethics approval statement:** NA.

**Patient consent statement:** NA.

**Permission to reproduce material from other sources:** NA.

**Clinical trial registration:** NA.

\blinenumbers

```{=tex}
\begin{center}
\textbf{Seabird assemblages are linked to the major western boundary current off eastern Australia}
\end{center}
```
*Running title:* Seabird assemblages off Eastern Australia

------------------------------------------------------------------------

**Abstract**

**Aim:** To identify assemblages of seabird species and relate them to environmental conditions off eastern Australia, a world seabird hotspot.

**Location:** East coast of Australia.

**Methods:** We used 15 seabird at-sea ship-based survey data sets collected over 37° of latitude, from 2016 to 2021, and fitted Region of Common Profile (RCP) mixture models for each season and type of data (presence-absence and abundance). RCP groups are defined as a set of regions where the probability of encountering a specific species profile is constant within regions, but different amongst them. These groups also vary according to covariates, which in our case included oceanographic, climatic, and physiographic parameters.

**Results:** Results were based on 142,646 seabirds recorded, from 80 species, including albatrosses, petrels,  prions, shearwaters, boobies, and terns, among other taxa. All models suggested two macro-scale assemblages (‘northern’ and ‘southern’), except for the autumn presence-absence model that identified three groups. The model results consistently show a biogeographic transition at \~34°S, near the latitude at which the East Australian Current (EAC) separates from the Australian continental slope. Sea surface temperatures or sea surface salinities were selected in all final models, further indicating a close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and abundance, resulted in similar spatial and species profile patterns.

**Main conclusions:** RCP models clearly identified two seabird assemblages off the east coast of Australia, suggesting the persistence of these groups at seasonal and macro spatial scales. Given ongoing poleward intensification that the EAC is experiencing, which is projected to continue over the next century, and its importance in influencing distributions of seabirds, the methods applied in our study could be replicated to assess possible changes in seabird assemblages, and how they are affected by changing environmental conditions.

***Keywords*** Assemblages; Biogeography; Community ecology; East Australian Current; Mixture models; Region of Common Profiles

@@

The EAC originates from the South Equatorial Current (SEC), a tropical nutrient-poor current. When the SEC's North Caledonian Jet meets the Queensland Plateau (\~18°S), it splits into a north-flowing branch towards the Gulf of Papua and a south-flowing branch, which marks the start of the EAC [@kessler2013; @ganachaud2014]. The South Caledonian Jet completes the EAC system at \~22°S after encountering the shelf break in the southern portion of the Great Barrier Reef (**Figure** \@ref(fig:fig1-study-area)**a**; @kessler2013; @ganachaud2014). The EAC interacts with the bathymetry of the seafloor, giving rise to oceanographic features and processes, such as smaller currents, upwellings, and persistent dynamic eddies [@oke2000; @ridgway2003a; @oke2019]. The EAC flows polewards and is primarily associated with the shelf break, but at \~32°S it detaches from the shelf and spawns an area of eddy generation that dissipates throughout the western Tasman Sea [@mata2006; @oke2019], the 'EAC eastern extension', often termed the 'Tasman Front' [@oke2019a]. Although most EAC energy is converted into eddies, the EAC still runs southwards following the shelf break towards Tasmania (the 'EAC southern extension'; **Figure** \@ref(fig:fig1-study-area)**a**), carrying eddy-like features as a transient current [@oke2019].

@@

Seabird data were gathered aboard *RV Investigator* and *RV Falkor* on 15 voyages, between 2016/17--2020/21, spanning over \~37° of latitude (**Figure** \@ref(fig:fig1-study-area)**b**; **Table S1**). One or more observers continuously recorded seabirds within a 300 m forward quadrant (eye height 25 m a.s.l) from sunrise to sunset hours while the ship was underway. The species, the numbers of birds, and their behaviours were logged in real time into a ship-based web server. Environmental parameters for atmospheric and oceanographic conditions were logged continuously by automated-ship based sensors, with seabird observations supplementing relevant environmental conditions such as visibility or associations with marine mammals. Individual records received a date/time/geographic coordinate label (multi-species records have the same label but are logged as separate entries for each species). Birds following the vessels were excluded from analyses. Seabird observation protocols followed @raymond2010, in use around Australia since 1980.

@@

Data wrangling, modelling and visualization were done in R 4.2.0 [@r2022]; the full list of packages can be found in the Supplementary Material. Code and the modelling dataset (see Section \ref{sec:modellingdatasets}) are available in **(Zenodo)**.

@@

Seabirds are highly mobile animals, and many species migrate and/or disperse during their non-breeding periods [@biology2002]. Therefore, seasonal differences in species composition were expected [e.g. @reid2023], and in the probabilities of occurrence and in total numbers. Thus, we built separate modelling datasets for each season. We aggregated seabird data by 1° latitude $\times$ 1° longitude grid cells (*c.* 110 $\times$ 110 km) and season (**Figure** \@ref(fig:fig1-study-area)**b**), by summing seabirds from the same species. To test if the presence-absence and abundance (raw counts) data would result in similar assemblage patterns, we built two species matrices, where the presence-absence matrix transformed values \>1 to 1. To avoid numerical instabilities in the models prompted by unusual or rare species, we kept only species occurring in at least six grids within each season (summer = 12 species; autumn = 21; winter = 20; spring = 33). The associated environmental data of each seabird record were averaged within the same grid cells and seasons, and environmental data were standardised before modelling.

@@

# References

::: {#refs}
:::

>>>>

author:

@@

@@

    affiliation: MarSci-Otago
    correspondingauthor: true

@@

    # email: eric.woehler@gmail.com
    affiliation: ASG

@@

    # email: matthew.schofield@otago.ac.nz
    affiliation: MatStat-Otago

@@

    # email: robert.smith@otago.ac.nz
    affiliation: MarSci-Otago

@@

    # email: lbugoni@yahoo.com.br
    affiliation: FURG

@@

    # email: will.rayment@otago.ac.nz
    affiliation: MarSci-Otago
address:
  - code: MarSci-Otago
    address: Department of Marine Science, University of Otago, Dunedin, Aotearoa New Zealand
  - code: ASG
    address: Australasian Seabird Group, BirdLife Australia, Hobart, Australia
  - code: MatStat-Otago
    address: Department of Mathematics and Statistics, University of Otago, Dunedin, Aotearoa New Zealand
  - code: FURG
    address: Institute of Biological Sciences, Universidade Federal do Rio Grande---FURG, Rio Grande, RS, Brazil

@@

  Identifying species assemblages helps understand the relationship between organisms and their environment. Assemblages can be used to predict biological changes caused by environmental perturbations, and are thus essential surrogates to monitor biodiversity. In this study, to identify and describe seabird assemblages, we used 15 at-sea ship-based survey data sets collected over 37° of latitude off eastern Australia, from 2016 to 2021. We fitted seasonal Region of Common Profile (RCP) mixture models, for two types of data (presence-absence and abundance). RCP groups are defined as regions where the probability of encountering a particular species profile is constant within regions, but different amongst them. These groups also vary according to covariates, which in our case included oceanographic, climatic, and physiographic parameters. Results were based on 142,646 seabirds recorded from 80 species, including albatrosses, petrels, prions, shearwaters, boobies, and terns, among other taxa. All models suggested two macro-scale assemblages (‘northern’ and ‘southern’), except for the autumn presence-absence model that identified three groups. The model results consistently show a biogeographic transition at \~34°S, near the latitude at which the East Australian Current (EAC) separates from the Australian continental slope. Sea surface temperatures or sea surface salinities were selected in all final models, further indicating a close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and abundance, resulted in similar spatial and species profile patterns. RCP models clearly identified two seabird assemblages off the east coast of Australia, suggesting the persistence of these groups at seasonal and macro spatial scales. Given the ongoing poleward intensification that the EAC is experiencing, which is projected to continue over the next century, and its importance in influencing the distributions of seabirds, the methods applied in our study could be replicated to assess possible changes in seabird assemblages and how they are affected by changing environmental conditions.

@@

journal: "Progress in Oceanography"
date: "2023-11-02"
classoption: preprint, 3p, authoryear
linenumbers: true
numbersections: true

@@

    base_format: rticles::elsevier_article
    keep_tex: true
    citation_package: default # Changing to 'default' allows specifying the CSL
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/elsevier-harvard.csl

@@

@@

  @hu2012

@@

```{r r-pckgs, include = FALSE, echo = FALSE}
library(readr)
library(dplyr)
library(purrr)
library(stringr)
library(kableExtra)
```

# Highlights {.unnumbered}

* The East Australian Current (EAC) is experiencing an ongoing poleward intensification
* We described seabird assemblages based on the Region of Common Profiles (RCP) method
* Assemblages of seabird species were correlated with the EAC
* Our analysis can be replicated to assess possible changes in seabird distribution
* The RCP method works well at the macro spatial scale for highly mobile animals

@@

The EAC originates from the South Equatorial Current (SEC), a tropical nutrient-poor current. When the SEC's North Caledonian Jet meets the Queensland Plateau (\~18°S), it splits into a north-flowing branch towards the Gulf of Papua and a south-flowing branch, which marks the start of the EAC [@kessler2013; @ganachaud2014]. The South Caledonian Jet completes the EAC system at \~22°S after encountering the shelf break in the southern portion of the Great Barrier Reef (**Fig.** \@ref(fig:fig1-study-area)**a**; @kessler2013; @ganachaud2014). The EAC interacts with the bathymetry of the seafloor, giving rise to oceanographic features and processes, such as smaller currents, upwellings, and persistent dynamic eddies [@oke2000; @ridgway2003a; @oke2019]. The EAC flows polewards and is primarily associated with the shelf break, but at \~32°S it detaches from the shelf and spawns an area of eddy generation that dissipates throughout the western Tasman Sea [@mata2006; @oke2019], the 'EAC eastern extension', often termed the 'Tasman Front' [@oke2019a]. Although most EAC energy is converted into eddies, the EAC still runs southwards following the shelf break towards Tasmania (the 'EAC southern extension'; **Fig.** \@ref(fig:fig1-study-area)**a**), carrying eddy-like features as a transient current [@oke2019].

```{r fig1-study-area, fig.cap = "\\label{fig1-study-area} Study area showing the main ocean currents (a), and seabird records made by season on top of the 1° latitude × 1° longitude grid cells (b). In (a), the East Australian Current (EAC) system is highlighted with its acronyms in bold and drawn in blue. The South Equatorial Current (SEC) and its branches, the North Caledonian Jet (NCJ) and South Caledonian Jet (SCJ), are drawn in red. The EAC 'eastern extension' is an eddy field (blue shaded area) often called the 'Tasman Front', from which the East Auckland Current originates (EAUC, in green). Figure (a) is adapted after Oke, Roughan, et al. (2019); for colour references see the online version", fig.align = "center", out.width = "100%", echo = FALSE}

@@

Seabird data were gathered aboard *RV Investigator* and *RV Falkor* on 15 voyages, between 2016/17--2020/21, spanning over \~37° of latitude (**Fig.** \@ref(fig:fig1-study-area)**b**; **Supplementary Table 1**). One or more observers continuously recorded seabirds within a 300 m forward quadrant (eye height 25 m a.s.l) from sunrise to sunset hours while the ship was underway. The species, the numbers of birds, and their behaviours were logged in real time into a ship-based web server. Environmental parameters for atmospheric and oceanographic conditions were logged continuously by automated-ship based sensors, with seabird observations supplementing relevant environmental conditions such as visibility or associations with marine mammals. Individual records received a date/time/geographic coordinate label (multi-species records have the same label but are logged as separate entries for each species). Birds following the vessels were excluded from analyses. Seabird observation protocols followed @raymond2010, in use around Australia since 1980.

@@

Data wrangling, modelling and visualization were done in R 4.2.0 [@r2022]; the full list of packages can be found in the Supplementary Material. Code and the modelling dataset (see Section \ref{sec:modellingdatasets}) are available in [@daudt_data_2023].

@@

Seabirds are highly mobile animals, and many species migrate and/or disperse during their non-breeding periods [@biology2002]. Therefore, seasonal differences in species composition were expected [e.g. @reid2023], and in the probabilities of occurrence and in total numbers. Thus, we built separate modelling datasets for each season. We aggregated seabird data by 1° latitude $\times$ 1° longitude grid cells (*c.* 110 $\times$ 110 km) and season (**Fig.** \@ref(fig:fig1-study-area)**b**), by summing seabirds from the same species. To test if the presence-absence and abundance (raw counts) data would result in similar assemblage patterns, we built two species matrices, where the presence-absence matrix transformed values \>1 to 1. To avoid numerical instabilities in the models prompted by unusual or rare species, we kept only species occurring in at least six grids within each season (summer = 12 species; autumn = 21; winter = 20; spring = 33). The associated environmental data of each seabird record were averaged within the same grid cells and seasons, and environmental data were standardised before modelling.

@@

# References {-}

<div id="refs"></div>

# Authors’ contribution (CRediT) {.unnumbered}

**Nicholas W. Daudt:** Conceptualization; Methodology; Software; Formal analysis; Data Curation; Visualization; Writing - Original Draft. **Eric J. Woehler:** Conceptualization; Methodology; Investigation; Resources; Data Curation; Writing - Review & Editing; Supervision. **Matthew R. Schofield:** Conceptualization; Software; Writing - Review & Editing; Supervision. **Robert O. Smith:** Writing - Review & Editing; Supervision. **Leandro Bugoni:** Writing - Review & Editing; Supervision. **William J. Rayment:** Conceptualization; Writing - Review & Editing; Supervision.
